A tool set for random number generation on GPUs in R

### 2021-12-24 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.tex
paper1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3058 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
paper1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3059 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
paper 1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3064 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\Abstract{
We introduce the \proglang{R} package \pkg{clrng} that leverages the \pkg{gpuR} package, and can generate random numbers on GPU by utilizing the \pkg{clRNG} (OpenCL) library. Parallel processing by graphics processing unit (GPU) can be used to speed up computationally intensive tasks, which when combined with \proglang{R}, it can largely improve \proglang{R}’s downsides in terms of slow speed, memory usage and computation mode.  There is currently no \proglang{R} package that does random number generation on GPU, while random number generation is critical in simulation-based statistical inference and modeling. \pkg{clrng} enables reproducible research by setting random streams on GPU and can thus accelerate several types of simulation and modelling. This package is portable and flexible, developers can use its random number generation kernel for various other purposes and applications. 
}

@@

Graphics Processing Units (GPUs) have the potential to make important contribution to parallel computing with \proglang{R}. GPUs can perform thousands of tests simultaneously, which makes them powerful for doing massively parallel computing, and they are relatively cheap compared to multicore CPU's. Although there have been a number of \proglang{R} packages developed which provide some GPU capability, they inevitably come with some limitations. Packages such as \pkg{gputools}, \pkg{gpumatrix}, \pkg{cudaBayesreg}, \pkg{rpud} (available on github), are now no longer maintained, the popular \pkg{tensorflow} \citep{tensorflow1} package uses GPU via \proglang{Python}, which makes it difficult to include as a dependency for new \proglang{R} packages. All of these mentioned packages are restricted to \proglang{R} users with NVIDIA GPUs. \pkg{gpuR} \citep{gpur1} is the only \proglang{R} package with a convenient and flexible interface between \proglang{R} and GPUs, and it is compatible with many GPU devices. By utilizing the \pkg{ViennaCL} \citep*{rupp2016viennacl} library, it provides a bridge between \proglang{R} and non-proprietary GPUs through the \proglang{OpenCL} (Open Computing Language) backend, which when combined with \pkg{Rcpp} \citep{rcpp1} gives a building block for other \proglang{R} packages.

@@

The \pkg{clrng} package described here is currently the only \proglang{R} package that is able to generate random numbers on GPUs. Accomplishing this is complicated to do because each process must produce an independent sequence of random numbers, and in order to ensure reproducibility it should be possible to save and restore the current state of each stream of numbers at any point.  Here we introduce the \proglang{R} package \pkg{clrng} that leverages the \pkg{gpuR} package, and can generate random numbers on GPU by using the \proglang{clRNG} \citep{l2015clrng} library, and can thus accelerate several types of statistical simulation and modelling.

The remaining sections are organized as follows:
In Section 2, we introduce streams and the use of streams in work-items on a GPU device for generating uniform random numbers, and the usage of \fct{clrng::runif}.
In Section 3, we introduce two non-uniform RNGs in \pkg{clrng}, as examples for users to develop other RNGs of interest on GPUs in \proglang{R}.
In Section 4, we apply GPU-generated uniform random numbers in Monte Carlo simulation for Fisher’s exact test, we introduce how the random numbers are used and how the algorithm is parallelized and implemented on GPU. Then we provide two real data examples to demonstrate the function usage and its \proglang{R} performance.
Section 5, we show an useful application of normal random numbers on GPU, using them to simulate batches of Gaussian random surfaces with Mat\'ern covariance matrices simultaneously, the simulation also uses GPU-enabled functions from our another package \pkg{gpuBatchMatrix}.
Finally, the paper concludes with a short summary and a discussion in Section 6.

@@

\proglang{clRNG} is an \proglang{OpenCL} library for uniform random number generation, it provides four different RNGs: the MRG31k3p, MRG32k3a, LFSR113, and Philox-4×32-10 generators. These four RNGs use different types of constructions. \pkg{clrng} package uses the MRG31k3p RNG, making it able to generate random numbers on GPUs. We choose MRG31k3p as the base generator for the following reasons:  the original \pkg{RNGStreams} package \citep{l2002object} was built with MRG32k3a, which was designed to be implemented in double precision, and not with 32-bit integers. The MRG31k3p generator was designed later, specifically for 32-bit integer arithmetic, so it runs faster on the 32-bit GPUs.  It is also faster than Philox-4×32-10 \citep{rLEC00b}. MRG31k3p was also statistically tested extensively and successfully, \cite[see][]{rLEC07b}.

@@

%THE PROBLEM:  each process needs to be independent, there's a risk of overlap.  solution: streams.  mrg31kp3 a stream is a vector of three positive integers.  initial state is first stream, compute new streams sequentially.
\subsection{Creating streams}\label{createstreams}
Random number generation is more complicated when the RNG algorithm is run in parallel processes in \proglang{R}, there is a risk that the generated sequences of random numbers have correlation. \pkg{clrng} uses multiple distinct streams that are used in work-items that executes in parallel on a GPU device. A popular way of obtaining multiple streams is to take an RNG with a long period and cut the RNG's sequence into very long disjoint pieces of equal length $Z$, and use each piece as a separate stream. Creating a new stream amounts to computing its starting point. In general, a stream object contains three elements: the current state of the stream, the initial state of the stream (or seed), and the initial state of the current substream (by default it is equal to the seed). Streams are created sequentially in the way that whenever the user creates a new stream, the software automatically jumps ahead by Z steps to find its initial state, and the three states in the stream object are set to it. Each of these streams can also be partitioned into substreams with equally-spaced starting points \citep{l2002object, rLEC15a}.

In \proglang{clRNG} library, the MRG31k3p RNG's entire period of length approximately $2^{185}$ is divided into approximately $2^{51}$ non-overlapping streams of length $Z = 2^{134}$. Each stream can be further partitioned into substreams of length $2^{72}$, although this is not currently implemented in \pkg{clrng}. The state (and seed) of each stream is a vector of six 31-bit integers. This size of state is appropriate for having streams running in work-items on GPU cards, while providing a sufficient period length for most applications. The initial state of the first stream (also called ``initial seed for the package'' in \proglang{clRNG} and ``initial'' in \pkg{clrng}) for the MRG31k3p is by default $(12345, 12345, 12345, 12345, 12345, 12345)$.

@@

Listing \ref{lst:normal} shows a fragment of the kernel that generates standard Gaussian random numbers. Aside from using local memory and the part that does the transformation, this kernel is the same with the kernel for generating uniform random numbers shown in Listing \ref{lst:uniformkernel}. If developers want to use other Gaussian transform methods, this kernel is easy to be adapted and incorporated in other \proglang{R} packages. The kernel executes over a 2-dimensional index space with $1 \times 2$ work-groups. \code{part[0]} and \code{part[1]} correspond to $R$ and $\Theta$ in the formulas respectively. \code{PI_2} is defined to be $\pi/2$. As the work-items in a work-group proceed at differing rates, \code{barrier(CLK_LOCAL_MEM_FENCE)} ensures correct ordering of memory operations to local memory, so that Gaussian random numbers $(X_1,Y_1), \dots, (X_n,Y_n)$ are generated in pairs correctly, errors such as $(X_n, Y_{(n-1)})$ or $(X_{(n-1)}, Y_{(n)})$ are avoided.
%

@@

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is above 160 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(n = 512 * 128)
R> system.time(clrng::rnorm(c(10000, 10000), streams = streams,
+    Nglobal = c(512, 128), type = "double"))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.051   0.000   0.052
\end{CodeOutput} 
\end{CodeChunk}

@@

Below is an example that produces a matrix of Exponential random numbers with expectation equal to 1.
\begin{CodeChunk}
\begin{CodeInput}
R> r_matrix <- clrng::rexp(c(2, 4), rate = 1, Nglobal = c(2,
+    2), type = "double")
R> as.matrix(r_matrix)
\end{CodeInput}
\begin{CodeOutput}
##       [,1]  [,2]  [,3]   [,4]
## [1,] 0.565 0.333 0.395 0.0932
## [2,] 1.535 0.866 0.124 0.0315
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(n = 256 * 64, initial = 666)
R> month_gpu <- vclMatrix(month, type = "integer")
R> system.time(result_month <- clrng::fisher.sim(month_gpu, 1000000,
+    streams = streams, type = "double", returnStatistics = TRUE,
+    Nglobal = c(256, 64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.319   0.003   0.321
\end{CodeOutput}
\begin{CodeInput}
R> result_month$threshold
\end{CodeInput}
\begin{CodeOutput}
## [1] -47955
\end{CodeOutput}
\begin{CodeInput}
R> result_month$simNum
\end{CodeInput}
\begin{CodeOutput}
## [1] 1015808
\end{CodeOutput}
\begin{CodeInput}
R> result_month$counts
\end{CodeInput}
\begin{CodeOutput}
## [1] 410825
\end{CodeOutput}
\begin{CodeInput}
R> result_month$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.404
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # using CPU
R> system.time(result_monthcpu <- stats::fisher.test(month, simulate.p.value = TRUE,
+    B = 1000000))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##  14.876   0.016  14.879
\end{CodeOutput}
\begin{CodeInput}
R> result_monthcpu$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.405
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> week_GPU <- gpuR::vclMatrix(week, type = "integer")
R> system.time(result_week <- clrng::fisher.sim(week_GPU, 10000000,
+    streams = streams, type = "double", returnStatistics = TRUE,
+    Nglobal = c(256, 64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##    1.93    0.02    1.95
\end{CodeOutput}
\begin{CodeInput}
R> result_week$threshold
\end{CodeInput}
\begin{CodeOutput}
## [1] -54990
\end{CodeOutput}
\begin{CodeInput}
R> result_week$simNum
\end{CodeInput}
\begin{CodeOutput}
## [1] 10010624
\end{CodeOutput}
\begin{CodeInput}
R> result_week$counts
\end{CodeInput}
\begin{CodeOutput}
## [1] 1193
\end{CodeOutput}
\begin{CodeInput}
R> result_week$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.000119
\end{CodeOutput} 
\end{CodeChunk}

The ``week'' table has a much smaller p-value: around 0.0001249672, which should require a larger number of simulations to get a more accurate p-value. With more than ten million simulations, we get 1205 cases and a p-value around 0.000120472 by \fct{fisher.sim}. \fct{stats::fisher.test} takes 91.6 seconds, while \fct{fisher.sim} takes about 4.282 seconds, the elapsed time is decreased to about 4.67\% of the time taken on CPU.

@@

\subsection{A summary of the results}
We summarized the comparison results in the table \ref{tab:summary} and plot the test statistics in Figure %\ref{fig4}.
\begin{table}

\caption{\label{tab:summarycompare}Summary of comparions of Fisher's test simulation on different devices. Computer 1 is equipped with CPU Intel Xenon W-2145 3.7Ghz and AMD Radeon VII. Computer 2 is equipped with VCPU Intel Xenon Skylake 2.5Ghz and VGPU Nvidia Tesla V100.\label{tab:summary}}
\centering
\begin{tabular}[t]{l|l|l|l|l|l}
\hline
\multicolumn{1}{c|}{ } & \multicolumn{2}{c|}{Computer 1} & \multicolumn{2}{c|}{Computer 2} & \multicolumn{1}{c}{ } \\
\cline{2-3} \cline{4-5}
B & Intel 2.5ghz & AMD Radeon & Intel 3.7ghz & NVIDIA V100 & Data\\
\hline
\multicolumn{6}{l}{\textbf{P-value}}\\
\hline
\hspace{1em}1M & 0.403804 & 0.403507 & 0.4035606 & 0.403507 & month\\
\hline
\hspace{1em}10M & 0.0001251 & 0.0001274 & 0.0001202 & 0.0001274 & week\\
\hline
\multicolumn{6}{l}{\textbf{Run-time}}\\
\hline
\hspace{1em}1M & 10.74 & 2.28 & 14.48 & 0.32 & month\\
\hline
\hspace{1em}10M & 63.24 & 10.82 & 91 & 1.95 & week\\
\hline
\end{tabular}
\end{table}

@@

\caption[Approximate sampling distributions of the test statistics from the two examples Month and Week]{Approximate sampling distributions of the test statistics from the two examples Month and Week. The test statistic values of the observed tables are marked with a blue line on each plot.}\label{fig:fighistMonth}
\end{figure}

\section{An application of normal RNG: Gaussian surface simulation} 
Given a parameter space $X$, a random field $U$ on $X$ is a collection of random variables $\{U_x :x\in X\}$. A Gaussian random field is a random field with the property that for any positive integer $n$ and any set of locations $x_1,\dots,x_n \in X$, the joint distribution of $U=(U_{x_1},\dots,U_{x_n})$ is multivariate Gaussian. The expectation of $U$ and covariance function $\Sigma$ completely determine the distribution of $U$, where 
\begin{gather*} 
U = [U(x_1), \ldots, U(x_n)]^\top \sim \text{MVN}(0, \Sigma) \\
\Sigma_{ij} = \cov[ U(x_i),U(x_j) ] = \sigma^2 \frac{2^{\kappa-1}}{\Gamma(\kappa)} \left(\sqrt{8\kappa} \frac{|x_i = x_j|}{\phi}\right)^\kappa  K_\kappa\left(\sqrt{8\kappa}  \frac{|x_i = x_j|}{\phi}\right).
\end{gather*}
A common choice for the covariance function $\Sigma$ is the Mat\'ern covariance \citep{matern1960spatial}. The Mat\'ern covariance between $U(x_i)$ and $U(x_j)$ takes the above form,
%$d =\|x-x'\|$ is the Euclidean distance between two spatial points, 
where $\sigma^2$ the variance of the random field $U$.
$\Gamma(\cdot)$ is the standard gamma function.  
$\phi$ is the range parameter or scale parameter with the dimension of distance, it controls the rate of decay of the correlation as distance increases. $K_\kappa(\cdot)$ is the modified Bessel function of the second kind with order $\kappa$, $\kappa$ is the shape parameter which determines the smoothness of $U(x)$, specifically, $U(x)$ is $m$ times mean-square differentiable if and only if $\kappa > m$. %(For $\kappa=0.5$, the Mat\'ern covariance function reduces to the exponential covariance function $\sigma^2*\exp(-d/\phi)$, when $\kappa \rightarrow \infty$, $\Sigma(d) \rightarrow \sigma^2*\exp{-(\|d\|/\phi)^2}$, which is called the Gaussian covariance.)  
There are several alternative parameterisations of Mat\'ern covariance functions in literatures  \citep[see][]{haskard2007anisotropic}. In \pkg{clrng}, we use the above form for computing Mat\'ern covariance.

%Due to the importance of Gaussian random field, there are a large amount of studies devoted to Gaussian random field generation techniques. 
\cite{LiuandLi2019} gives a comprehensive review on 7 popular methods for Gaussian random field generation, %which are the turning bands method \citep{matheron_1973}, spectral method \citep{Meja1974OnTS,1972JSV}, matrix decomposition method \citep{davis1987}, Karhunen-Lo\'eve expansion \citep{LoeveM:1978}, moving average method \citep{journel1974, oliver1995moving}, sequential simulation \citep{johnson1987multivariate, gomez1993joint, pebesma2004multivariable}, and local average subdivision \citep{fenton1990simulation}. 
all of these methods except matrix decomposition method, are approximations and has specific requirements on the type of grid or covariance functions. The matrix decomposition method is exact, it works for all covariance functions and can generate random field on all types of grids. Other \proglang{R} packages that offer simulation of Gaussian random fields such as \pkg{geoR} \citep{geoR2001}, does not work for large number of locations; the \pkg{RandomFields} \citep{RandomFields2015,RandomFields2020} package can use different methods for simulation of Gaussian fields, among which the modified circulant embedding method \citep{Dietrich1997FastAE} for covariance matrix decomposition is also an exact method, however, it works only for isotropic Gaussian fields and on rectangular grids. \pkg{clrng} does simulation of exact Gaussian fields on the GPU as it relies on the matrix decomposition method. The implementation of this method is as follows.

Suppose $U=(U_1, U_2, \dots, U_n)$ is a Gaussian random field with mean $\mu$ and covariance matrix $\Sigma$, without loss of generality, we assume mean value zero $(\mu = 0)$. Since covariance matrix $\Sigma$ is symmetric and positive-definite, we can take Cholesky decomposition of $\Sigma$, then we can simulate random samples from $U$ using Algorithm \ref{algorithm2},

>>>>

\Abstract{
We introduce the \proglang{R} package \pkg{clrng} that leverages the \pkg{gpuR} package, and can generate random numbers on GPU by utilizing the \pkg{clRNG} (OpenCL) library. Parallel processing by graphics processing unit (GPU) can be used to speed up computationally intensive tasks, which when combined with \proglang{R}, it can largely improve \proglang{R}’s downsides in terms of slow speed, memory usage and computation mode.  %There is currently no \proglang{R} package that does random number generation on GPU, while random number generation is critical in simulation-based statistical inference and modeling. 
\pkg{clrng} enables reproducible research by setting random streams on GPU and can thus accelerate several types of simulation and modelling. This package is portable and flexible, developers can use its random number generation kernel for various other purposes and applications. 
}

@@

Graphics Processing Units (GPUs) have the potential to make an important contribution to parallel computing with \proglang{R}. GPUs can perform thousands of computations simultaneously, which makes them powerful for doing massively parallel computing, and they are relatively cheap compared to multicore CPU's. Although there have been a number of \proglang{R} packages developed which provide some GPU capability, they inevitably come with some limitations. Packages such as \pkg{gputools}, \pkg{gpumatrix}, \pkg{cudaBayesreg}, \pkg{rpud} (available on github), are no longer maintained, the popular \pkg{tensorflow} \citep{tensorflow1} package uses GPU via \proglang{Python}, which makes it difficult to include as a dependency for new \proglang{R} packages. All of these mentioned packages are restricted to \proglang{R} users with NVIDIA GPUs. \pkg{gpuR} \citep{gpur1} is the only \proglang{R} package with a convenient and flexible interface between \proglang{R} and GPUs, and it is compatible with many GPU devices. By utilizing the \pkg{ViennaCL} \citep*{rupp2016viennacl} library, it provides a bridge between \proglang{R} and non-proprietary GPUs through the \proglang{OpenCL} (Open Computing Language) backend, which when combined with \pkg{Rcpp} \citep{rcpp1} gives a building block for other \proglang{R} packages.

@@

The \pkg{clrng} package described here is currently the only \proglang{R} package that is able to generate random numbers on GPUs. Accomplishing this is complicated because each process must produce an independent sequence of random numbers, and in order to ensure reproducibility it should be possible to save and restore the current state of each stream of numbers at any point.  Here we introduce the \proglang{R} package \pkg{clrng} that leverages the \pkg{gpuR} package, and can generate random numbers on GPU by using the \proglang{clRNG} \citep{l2015clrng} library, and can thus accelerate several types of statistical simulation and modelling.

The remaining sections are organized as follows:
In Section 2, we introduce streams and the use of streams in work-items on a GPU device for generating uniform random numbers, and the usage of \fct{clrng::runif}.
In Section 3, we introduce two non-uniform RNGs in \pkg{clrng}, as examples for users to develop other RNGs of interest on GPUs in \proglang{R}.
In Section 4, we apply GPU-generated uniform random numbers in Monte Carlo simulation for Fisher’s exact test, we introduce how the random numbers are used and how the algorithm is parallelized and implemented on GPU. Then we provide two real data examples to demonstrate the function usage and its \proglang{R} performance.
In Section 5, we show a useful application of normal random numbers on GPU, using them to simulate batches of Gaussian random surfaces with Mat\'ern covariance matrices simultaneously, the simulation also uses GPU-enabled functions from our other package \pkg{gpuBatchMatrix}.
Finally, the paper concludes with a short summary and a discussion in Section 6.

@@

\proglang{clRNG} is an \proglang{OpenCL} library for uniform random number generation, it provides four different RNGs: the MRG31k3p, MRG32k3a, LFSR113, and Philox-4×32-10 generators. These four RNGs use different types of constructions. The \pkg{clrng} package uses the MRG31k3p RNG, making it able to generate random numbers on GPUs. We choose MRG31k3p as the base generator for the following reasons:  the original \pkg{RNGStreams} package \citep{l2002object} was built with MRG32k3a, which was designed to be implemented in double precision, and not with 32-bit integers. The MRG31k3p generator was designed later, specifically for 32-bit integer arithmetic, so it runs faster on the 32-bit GPUs. The MRG31k3p generator was introduced in the paper \citep{rLEC00b}. It is also faster than Philox-4×32-10. MRG31k3p was also statistically tested extensively and successfully, \cite[see][]{rLEC07b}.

@@

%THE PROBLEM:  each process needs to be independent, there's a risk of overlap.  solution: streams.  mrg31kp3 a stream is a vector of three positive integers.  initial state is first stream, compute new streams sequentially.
\subsection{Creating streams}\label{createstreams}
Random number generation is more complicated when the RNG algorithm is run in parallel processes in \proglang{R}, there is a risk that the generated sequences of random numbers have correlation. \pkg{clrng} uses multiple distinct streams that are used in work-items that executes in parallel on a GPU device. A popular way of obtaining multiple streams is to take an RNG with a long period and cut the RNG's sequence into very long disjoint pieces of equal length $Z$, and use each piece as a separate stream. Creating a new stream amounts to computing its starting point. Each of these streams can also be partitioned into substreams with equally-spaced starting points \citep{l2002object, rLEC15a}, although this is not currently implemented in \pkg{clrng}.. In general, a stream object contains three elements: the current state of the stream, the initial state of the stream (or seed), and the initial state of the current substream (by default it is equal to the seed). Streams are created sequentially in the way that whenever the user creates a new stream, the software automatically jumps ahead by Z steps to find its initial state, and the three states in the stream object are set to it.

In the \proglang{clRNG} library, the MRG31k3p RNG's entire period of length approximately $2^{185}$ is divided into approximately $2^{51}$ non-overlapping streams of length $Z = 2^{134}$. %Each stream can be further partitioned into substreams of length $2^{72}$. 
The state (and seed) of each stream is a vector of six 31-bit integers. This size of state is appropriate for having streams running in work-items on GPU cards, while providing a sufficient period length for most applications. The initial state of the first stream (also called ``initial seed for the package'' in \proglang{clRNG} and ``initial'' in \pkg{clrng}) for the MRG31k3p is by default $(12345, 12345, 12345, 12345, 12345, 12345)$.

@@

Listing \ref{lst:normal} shows a fragment of the kernel that generates standard Gaussian random numbers. Aside from using local memory and the part that does the Box-Muller transformation, this kernel is the same with the kernel for generating uniform random numbers shown in Listing \ref{lst:uniformkernel}. If developers want to use other Gaussian transform methods, this kernel is easy to be adapted and incorporated in other \proglang{R} packages. This kernel executes over a 2-dimensional index space with $1 \times 2$ work-groups. \code{part[0]} and \code{part[1]} correspond to $R$ and $\Theta$ in the formulas respectively. \code{PI_2} is defined to be $\pi/2$. As the work-items in a work-group proceed at different rates, \code{barrier(CLK_LOCAL_MEM_FENCE)} ensures correct ordering of memory operations to local memory, so that Gaussian random numbers $(X_1,Y_1), \dots, (X_n,Y_n)$ are generated in pairs correctly, errors such as $(X_n, Y_{(n-1)})$ or $(X_{(n-1)}, Y_{(n)})$ are avoided.
%

@@

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is above 160 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(n = 512 * 128)
R> system.time(clrng::rnorm(c(10000, 10000), streams = streams,
+    Nglobal = c(512, 128), type = "double"))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.047   0.000   0.048
\end{CodeOutput} 
\end{CodeChunk}

@@

Below is an example that produces a $2 \times 4$ matrix of Exponential random numbers with expectation equal to 1.
\begin{CodeChunk}
\begin{CodeInput}
R> r_matrix <- clrng::rexp(c(2, 4), rate = 1, Nglobal = c(2,
+    2), type = "double")
R> as.matrix(r_matrix)
\end{CodeInput}
\begin{CodeOutput}
##       [,1]  [,2]  [,3]   [,4]
## [1,] 0.565 0.333 0.395 0.0932
## [2,] 1.535 0.866 0.124 0.0315
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(n = 256 * 64, initial = 666)
R> month_gpu <- vclMatrix(month, type = "integer")
R> system.time(result_month <- clrng::fisher.sim(month_gpu, 1000000,
+    streams = streams, type = "double", returnStatistics = TRUE,
+    Nglobal = c(256, 64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.311   0.004   0.315
\end{CodeOutput}
\begin{CodeInput}
R> result_month$threshold
\end{CodeInput}
\begin{CodeOutput}
## [1] -47955
\end{CodeOutput}
\begin{CodeInput}
R> result_month$simNum
\end{CodeInput}
\begin{CodeOutput}
## [1] 1015808
\end{CodeOutput}
\begin{CodeInput}
R> result_month$counts
\end{CodeInput}
\begin{CodeOutput}
## [1] 410825
\end{CodeOutput}
\begin{CodeInput}
R> result_month$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.404
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # using CPU
R> system.time(result_monthcpu <- stats::fisher.test(month, simulate.p.value = TRUE,
+    B = 1015808))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##  14.791   0.027  14.826
\end{CodeOutput}
\begin{CodeInput}
R> result_monthcpu$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.405
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> week_GPU <- gpuR::vclMatrix(week, type = "integer")
R> system.time(result_week <- clrng::fisher.sim(week_GPU, 10000000,
+    streams = streams, type = "double", returnStatistics = TRUE,
+    Nglobal = c(256, 64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   1.944   0.021   1.963
\end{CodeOutput}
\begin{CodeInput}
R> result_week$threshold
\end{CodeInput}
\begin{CodeOutput}
## [1] -54990
\end{CodeOutput}
\begin{CodeInput}
R> result_week$simNum
\end{CodeInput}
\begin{CodeOutput}
## [1] 10010624
\end{CodeOutput}
\begin{CodeInput}
R> result_week$counts
\end{CodeInput}
\begin{CodeOutput}
## [1] 1193
\end{CodeOutput}
\begin{CodeInput}
R> result_week$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.000119
\end{CodeOutput} 
\end{CodeChunk}

The ``week'' table has a much smaller p-value: around 0.000125, which should require a larger number of simulations to get a more accurate p-value. With more than ten million simulations, we get 1205 cases and a p-value around 0.000120472 by \fct{fisher.sim}. \fct{stats::fisher.test} takes 92.45 seconds, while \fct{fisher.sim} takes about 1.95 seconds, the GPU elapsed time is decreased to about 2.10\% of the time taken on CPU.

@@

\subsection{A summary of the results}
We summarized the comparison results in Table \ref{tab:summary} and plot the test statistics in Figure \ref{fig4}.
\begin{table}

\caption{\label{tab:summarycompare}Summary of comparions of Fisher's test simulation on different devices. Computer 1 is equipped with CPU Intel Xenon W-2145 3.7Ghz and AMD Radeon VII. Computer 2 is equipped with VCPU Intel Xenon Skylake 2.5Ghz and VGPU Nvidia Tesla V100.\label{tab:summary}}
\centering
\begin{tabular}[t]{l|l|l|l|l|l}
\hline
\multicolumn{1}{c|}{ } & \multicolumn{2}{c|}{Computer 1} & \multicolumn{2}{c|}{Computer 2} & \multicolumn{1}{c}{ } \\
\cline{2-3} \cline{4-5}
B & Intel 2.5ghz & AMD Radeon & Intel 3.7ghz & NVIDIA V100 & Data\\
\hline
\multicolumn{6}{l}{\textbf{P-value}}\\
\hline
\hspace{1em}1M & 0.403804 & 0.403507 & 0.4035606 & 0.403507 & month\\
\hline
\hspace{1em}10M & 0.0001251 & 0.0001274 & 0.0001202 & 0.0001274 & week\\
\hline
\multicolumn{6}{l}{\textbf{Run-time}}\\
\hline
\hspace{1em}1M & 10.74 & 2.28 & 14.8 & 0.32 & month\\
\hline
\hspace{1em}10M & 63.24 & 10.82 & 92 & 1.95 & week\\
\hline
\end{tabular}
\end{table}

@@

\caption[\label{fig4}Approximate sampling distributions of the test statistics from the two examples Month and Week]{\label{fig4}Approximate sampling distributions of the test statistics from the two examples Month and Week. The test statistic values of the observed tables are marked with a blue vertical line on each plot.}\label{fig:fighistMonth}
\end{figure}

\section{An application of normal RNG: Gaussian surface simulation} 
Given a parameter space $X$, a random field $U$ on $X$ is a collection of random variables $\{U_x :x\in X\}$. A Gaussian random field is a random field with the property that for any positive integer $n$ and any set of locations $x_1,\dots,x_n \in X$, the joint distribution of $U=(U_{x_1},\dots,U_{x_n})$ is multivariate Gaussian. The expectation of $U$ and covariance function $\Sigma$ completely determine the distribution of $U$, where 
\begin{gather*} 
U = [U(x_1), \ldots, U(x_n)]^\top \sim \text{MVN}(0, \Sigma), \\
\Sigma_{ij} = \cov[ U(x_i),U(x_j) ] = \sigma^2 \frac{2^{\kappa-1}}{\Gamma(\kappa)} \left(\sqrt{8\kappa} \frac{|x_i = x_j|}{\phi}\right)^\kappa  K_\kappa\left(\sqrt{8\kappa}  \frac{|x_i = x_j|}{\phi}\right),
\end{gather*}
where \text{MVN} standards for the multivariate normal distribution, and $\Sigma$ is the covariance matrix. A common choice for $\Sigma$ is the Mat\'ern covariance \citep{matern1960spatial}. The Mat\'ern covariance between $U(x_i)$ and $U(x_j)$ takes the above form,
%$d =\|x-x'\|$ is the Euclidean distance between two spatial points, 
where $\sigma^2$ the variance of the random field $U$.
$\Gamma(\cdot)$ is the standard gamma function.  
$\phi$ is the range parameter or scale parameter with the dimension of distance, it controls the rate of decay of the correlation as distance increases. $K_\kappa(\cdot)$ is the modified Bessel function of the second kind with order $\kappa$, $\kappa$ is the shape parameter which determines the smoothness of $U(x)$, specifically, $U(x)$ is $m$ times mean-square differentiable if and only if $\kappa > m$. %(For $\kappa=0.5$, the Mat\'ern covariance function reduces to the exponential covariance function $\sigma^2*\exp(-d/\phi)$, when $\kappa \rightarrow \infty$, $\Sigma(d) \rightarrow \sigma^2*\exp{-(\|d\|/\phi)^2}$, which is called the Gaussian covariance.)  
There are several alternative parameterisations of Mat\'ern covariance functions in literature  \citep[see][]{haskard2007anisotropic}. In \pkg{clrng}, we use the above form for computing Mat\'ern covariance.

%Due to the importance of Gaussian random field, there are a large amount of studies devoted to Gaussian random field generation techniques. 
\cite{LiuandLi2019} gives a comprehensive review on seven popular methods for Gaussian random field generation, %which are the turning bands method \citep{matheron_1973}, spectral method \citep{Meja1974OnTS,1972JSV}, matrix decomposition method \citep{davis1987}, Karhunen-Lo\'eve expansion \citep{LoeveM:1978}, moving average method \citep{journel1974, oliver1995moving}, sequential simulation \citep{johnson1987multivariate, gomez1993joint, pebesma2004multivariable}, and local average subdivision \citep{fenton1990simulation}. 
all of these methods except matrix decomposition method, are approximations and has specific requirements on the type of grid or covariance functions. The matrix decomposition method is exact, it works for all covariance functions and can generate random field on all types of grids. Other \proglang{R} packages that offer simulation of Gaussian random fields such as \pkg{geoR} \citep{geoR2001}, does not work for large number of locations; the \pkg{RandomFields} \citep{RandomFields2015,RandomFields2020} package can use different methods for simulation of Gaussian fields, among which the modified circulant embedding method \citep{Dietrich1997FastAE} for covariance matrix decomposition is also an exact method, however, it works only for isotropic Gaussian fields and on rectangular grids. \pkg{clrng} does simulation of exact Gaussian fields on the GPU as it relies on the matrix decomposition method. The implementation of this method is as follows.

Suppose $U=(U_1, U_2, \dots, U_n)^\top$ is a Gaussian random field with mean $\mu$ and covariance matrix $\Sigma$, without loss of generality, we assume mean value zero $(\mu = 0)$. Since covariance matrix $\Sigma$ is symmetric and positive-definite, we can take Cholesky decomposition of $\Sigma$, then we can simulate random samples from $U$ using Algorithm \ref{algorithm2},

\begin{algorithm}[H]
\SetAlgoLined
 1, Calculate the covariance matrix $\Sigma$ between locations\;
 2, Compute the Cholesky decomposition of $\Sigma = L \cdot D \cdot L^\top$, where $L$ is a lower unit triangular matrix, and $D$ is a diagonal matrix\;
 3, Generate on GPU a random matrix $Z=(Z_1, Z_2, \dots, Z_n) \sim \text{MVN}(0,I_n)$, where $I_n$ is a $n \times n$ identity matrix\;
 4, Compute the random samples from $U$ in batches using $U = L \cdot D^{\frac{1}{2}} \cdot Z$ \;
 \caption{Gaussian random fields simulation using covariance matrix decomposition method.}
 \label{algorithm2}
\end{algorithm}
